---
title: "Mid-Project Group 3"
output: html_notebook
---
```{r}
# Read CSV from a URL

# Import the dataset from Google Drive
url <- "https://drive.google.com/uc?export=download&id=1d8bEgUisDBY-2qC9Csy4qm3d_th_l9PN"
df <- read.csv(url, na.strings = c("NA", "", "NULL"))

# View the first few rows of the dataframe to check the import
head(df)


```
```{r}
#View the summary of the dataset
summary(df)
```
## Data Exploration
Univariate Analysis


HISTOGRAM (NUMERIC)

```{r}
# Histogram of Screen time hours
hist(df$screen_time_hours, 
     main="Histogram of screen time hour", 
     col="cyan", 
     xlab="Screen time hour", 
     border="black")
```

Observation: Most users had screen time between 3 and 8 hours. Few users had very high usage above 10 hours.

```{r}
# Histogram of   bedtime_hour
hist(df$  bedtime_hour, 
     main="Histogram of bedtime hour", 
     col="cyan", 
     xlab="Bedtime Hour", 
     border="black")
```
Observation: Most people sleep 6–8 hours, some less than 5 or more than 9 

```{r}
# Histogram of age 
hist(df$   age , 
     main="Histogram of age ", 
     col="cyan", 
     xlab="Age", 
     border="black")

```

the most common users are around age of 18


```{r}
 # Density Plot
plot (density(df$  screen_time_hours), 
     main="Density Plot of screen_time_hours", 
     col="GREEN", 
     lwd=3)
```
peaks around 5–7 hours

```{r}
# Density Plot
plot(density(df$bedtime_hour), 
     main="Density Plot of bedtime_hour", 
     col="BLUE", 
     lwd=3)
```
Peak around 22:00 

```{r}
# Density Plot
plot(density(df$age), 
     main="Density Plot of age", 
     col="green", 
     lwd=3)
#
```
Most common user are between age of 20 to 25


```{r}
# Boxplot
boxplot(df$age, 
        main="Box Plot of age", 
        col="cyan")
```
Mostly normal, no outliers spotted

```{r}
# Boxplot
boxplot(df$screen_time_hours, 
        main="Box Plot of screen_time_hours", 
        col="cyan")
```
Outliers exist (high screen time)

```{r}
boxplot(df$bedtime_hour, 
        main="Box Plot of bedtime_hour", 
        col="cyan")
```
Mostly normal, no outliers spotted

Exploring Categorical Data
```{r}
 #Count 
table(df$app_type_most_used)
```

```{r}
#Count of Home Country
table(df$stress_level)
```


```{r}
# Bar plot
barplot(table(df$app_type_most_used), 
        main="Count of app_type_most_used", 
        col=c("cyan", "green"))
```
#Educational apps are most used then gaming and utility respectively

```{r}
# Bar plot 
barplot(table(df$stress_level), 
        main="Count of stress_level", 
        col=c("cyan","red", "green"))

```
Stress level moderate is highly distributed 



Multivariate plots

Scatter plot, pair plot and correlation matrix have been displayed to observe different result

```{r}
plot(df$age, df$screen_time_hours,
     main = "Scatter Plot of Age vs Screen Time",
     xlab = "Age",
     ylab = "Screen Time (Hours)",
     col = "blue",
     pch = 16)
```

outliers spotted here


```{r}
plot(df$age, df$bedtime_hour,
     main = "Scatter Plot of Age vs Bedtime Hour",
     xlab = "Age",
     ylab = "Bedtime Hour",
     col = "darkgreen",
     pch = 16)
```


```{r}
plot(df$screen_time_hours, df$bedtime_hour,
     main = "Scatter Plot of Screen Time vs Bedtime Hour",
     xlab = "Screen Time (Hours)",
     ylab = "Bedtime Hour",
     col = "purple",
     pch = 16)

```
```{r}
# Colored pair plot by User_Type
# Load the required package
library(GGally)

# Subset relevant numeric columns
df_subset <- df[, c("age", "screen_time_hours", "bedtime_hour")]

# Create the pair plot
ggpairs(df_subset,
        title = "Pair Plot: Age, Screen Time, and Bedtime Hour",
        upper = list(continuous = wrap("cor", size = 4)),
        lower = list(continuous = wrap("points", alpha = 0.7)),
        diag = list(continuous = wrap("densityDiag")))


```

More screen time is basically high bed time hour

Correlation (Numeric vs. Numeric)
```{r}
cor(df$screen_time_hours, df$bedtime_hour

)
```
```{r}
cor(df[,1:2])
```
Box Plot (Numeric vs. Categorical)
```{r}
boxplot(bedtime_hour
 ~ stress_level, data=df,
        main="Box Plot of Bed time hour by stress label",
        col=c("green", "red"))
```

```{r}
boxplot(screen_time_hours ~ stress_level, data=df,
        main="Box Plot of Screen time  hour by stress_level",
        col=c("green", "red"))
```
In this Box plot more outliers were spotted in moderate stress level.


##Check for missing values
```{r}
is.na(df)
```
Count missing values per column
```{r}
colSums(is.na(df))
```
Check total missing values
```{r}
sum(is.na(df))
```
250 missing values observed
Replace NA with the Median of the Column 1,2,3
```{r}
# Replace NA with the Median of each column
df$screen_time_hours[is.na(df$screen_time_hours)] <- median(df$screen_time_hours, na.rm = TRUE)
df$bedtime_hour[is.na(df$bedtime_hour)] <- median(df$bedtime_hour, na.rm = TRUE)
df$age[is.na(df$age)] <- median(df$age, na.rm = TRUE)

# Print the updated dataframe
print(df[c("screen_time_hours", "bedtime_hour", "age")])


```

Replace Categorical column app_type_most_used NA with Mode
```{r}
# Replace missing values in 'app_type_most_used' with the mode
df$app_type_most_used[is.na(df$app_type_most_used)] <- 
  names(sort(table(df$app_type_most_used), decreasing = TRUE))[1]

# Print only the 'app_type_most_used' column
print(df[c("app_type_most_used")])
```

Remove rows of missing values in Target column stress_level

```{r}
# Remove rows where 'stress_level' is NA
df <- subset(df, !is.na(stress_level))


# Print the 'stress_level' column as part of the data frame
print(df[c("stress_level")])


```


##Data type conversion using label coding

```{r}
# Apply label encoding to 'app_type_most_used'
df$app_type_most_used_encoded <- as.integer(factor(df$app_type_most_used, 
                                                   levels = c("educational", "entertainment", "social", "gaming", "utility")))

# Apply label encoding to 'stress_level'
df$stress_level_encoded <- as.integer(factor(df$stress_level, 
                                             levels = c("low", "moderate", "high")))

# Print only the two encoded columns as a dataframe
print(df[c("app_type_most_used_encoded", "stress_level_encoded")])




```




##Outlier detection and capping using IQR method


```{r}
# Function to detect outliers using IQR
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  return(x < lower_bound | x > upper_bound)
}

# Detect outliers for each column
outliers_screen_time <- detect_outliers(df$screen_time_hours)
outliers_bedtime <- detect_outliers(df$bedtime_hour)
outliers_age <- detect_outliers(df$age)

# Count the number of outliers for each column
outliers_screen_time_count <- sum(outliers_screen_time)
outliers_bedtime_count <- sum(outliers_bedtime)
outliers_age_count <- sum(outliers_age)

# Print the count of outliers in each column
print(paste("Number of outliers in screen_time_hours: ", outliers_screen_time_count))
print(paste("Number of outliers in bedtime_hour: ", outliers_bedtime_count))
print(paste("Number of outliers in age: ", outliers_age_count))

df$screen_time_hours[df$screen_time_hours < lower_bound] <- lower_bound
df$screen_time_hours[df$screen_time_hours > upper_bound] <- upper_bound

print(df$screen_time_hours[outliers_screen_time])
```




##Data transformation (scaling)


```{r}
# Scale the numerical columns: screen_time_hours, bedtime_hour, and age
df$screen_time_hours_scaled <- scale(df$screen_time_hours)
df$bedtime_hour_scaled <- scale(df$bedtime_hour)
df$age_scaled <- scale(df$age)

# Print the first few rows of the scaled data
print(df[, c("screen_time_hours_scaled", "bedtime_hour_scaled", "age_scaled")])

```
##New Dataset after preprocessing 
```{r}
# Remove the specified columns from the dataframe
df_new <- df[, !names(df) %in% c("screen_time_hours", "bedtime_hour", "age", "screen_time_hours_capped", 
                                 "app_type_most_used", "stress_level")]

# Print the cleaned dataframe
print(df_new)



```
```{r}
# Check for missing values in the final dataset (df)
missing_values <- sum(is.na(df))

# Print the result
if(missing_values > 0) {
  cat("There are", missing_values, "missing values in the dataset.\n")
} else {
  cat("No missing values in the dataset.\n")
}

```
##Recheked again and ensured My dataset has no missing values. All coloumns are properly encoded and removed outliers and scaled.



```

